---
title: "Take-home Exercise 3 Part 1"
author: "Lai Meihao"
date: "March 25, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1.Overview

## 1.1 Project Brief

Take home exercise 3 is a preliminary work of the group project - blue pulse. Water is essential to life, yet billions worldwide still lack access to safe drinking water and sanitation. Waterborne disease such as acute hepatitis A and diarrhea continue to pose series public health threats, particularly in regions with inadequate water access. SDG 6 aims to ensure universal access to water and sanitation The project aims to understand the 1) understand the distribution of water access and safety by region, 2) analyze the long-term effects of water access on health outcome.

The project team consists of three members and each member will take one of the main prototype modules as follows:

-   Exploratory Data Analysis (initial & treemap)

-   Panel Linear Model for observations over time for multiple entities (regions, countries)

-   Confirmatory Data Analysis.

## 1.2 Project Objectives

The project will using WDI from world bank and burden of disease from institute for health metrics and evaluation. The objective of this take-home exercise is to build a prototype - user interface design on Panel Linear Model.

# 2. Initial Data Preparation

## 2.1 Install and launch R packages

The project uses `p_load()` of pacman package to check if the R package are installed in the computer.

The following code chunk is used to install and launch the R packages.

```{r}
pacman::p_load(tidyverse, ggplot2, plm,ExPanDaR,lubridate, ggthemes, plotly,dplyr,purrr,forecast)
```

## 2.2 Import Data

The project examine the dataset from WDI from world bank and burden of disease from institute for health metrics and evaluation focusing on WHO region countries from 2000 to 2021.\
The data have already have been initally processed and merged by member of the team - Xinyue.

The below code chunk use `read_csv()` to read the csv file into R as a data frame

```{r}
merged_data <- read_csv("data/merged_data.csv")

```

## **2.3 Overview of the data**

### **Dataset Structure**

The below code chunk check the structure of the data.

```{r}
# Preview structure and types
glimpse(merged_data)

# View first few rows
head(merged_data)

# View in spreadsheet-style window
View(merged_data)

```

# 3.Data Processing

## 3.1 Missing Data

The below code chunk of `is.na()` to check the missing values in the `merged_data` frame, and returns a logical matrix( or vector, depending on the data structure. `sum()` function calculate the sum of the elements of the logical matrix or vector.

```{r}
sum(is.na(merged_data))

```

The result of zero indicate that there no missing in the merge dataset.

## 3.2 Duplicate Value

The `duplicated()` function checks for duplicate rows in the `merged_data` frame.

```{r}
sum(duplicated(merged_data))
```

The result indicate that there is no duplicate in the `merged_data`.

# 4.PLM: Linear Model for panel data

## 4.1 Overview of plm

A **Panel Linear Model (PLM)** is a statistical model used to analyze data that includes observations over time (time-series) and across different units (cross-sectional). This type of data, often called panel data or longitudinal data, involves multiple subjects (such as countries, individuals, or companies) observed over several time periods.

For this project, we are using data from the World Development Indicators (WDI) as WASH (Water, Sanitation, and Hygiene) indicators and data from the Institute for Health Metrics as health indicators. We will apply the PLM to estimate the coefficients of the WASH indicators for each health indicator. Subsequently, we will use time-series analysis to forecast trends for the health indicators, particularly disease rates, based on the coefficients obtained from the WASH indicators.

To apply PLM in our project, we will be using [`plm`](https://cran.r-project.org/web/packages/plm/plm.pdf) package in R.

## 4.2 Creating Panel Data

Before proceed to the PLM analysis, we will first create panel data. We will be start by understanding `merged_data`.\
The below code chunk use `names()` function to extract the names of the columns in the `merged_data`.

```{r}
names(merged_data)

```

The output shows the columns in the `merged_data`.

The below code chunk use the `summary()` function to provide a quick overview of each column in a dataframe.

```{r}
summary(merged_data)
```

This summary provides key insights into the distribution and spread of values in each of the columns. The `AvgValue` column, for instance, has a wide range, with very large maximum values compared to the median and first quartile. Similarly, health indicators like `TyphoidRate`, `DiarrheaRate`, and `HepatitisRate` show varying levels of distribution, with some regions having very low rates and others experiencing high rates.

The below code chunk use `str()` function to display the structure of the merged_data. It provides a compact overview of the data type of each column, as well as the first few entries in each column.

```{r}
str(merged_data)

```

The dataset is a **spec_tbl_df** (which is a specific type of tibble in R) with **1232 rows and 8 columns**. The dataset includes 8 variables (columns), and it uses the `spec_tbl_df` class from the `tibble` package.

The below code chunk summaries the `merged_data` set by grouping it by `Region` and `Year`.

The result of this code is a new data frame, `merged_data_clean`, where:

-   Each row corresponds to a unique combination of `Region` and `Year`.

-   For each `Region` and `Year`, the mean values for `TyphoidRate`, `DiarrheaRate`, `HepatitisRate`, and `UnsafeRisk` are calculated, ignoring missing values.

The code chunk below use the `pivot_wider()` function from the `tidyverse` package to transform the `merged_data` dataset from long format to wide format. As the WASH indicator are store in the column `Series.Name`.

```{r}
wide_data <- merged_data %>%
  pivot_wider(
    id_cols = c(Region, Year, TyphoidRate, DiarrheaRate, HepatitisRate, UnsafeRisk),
    names_from = Series.Name,
    values_from = AvgValue
  )
```

The code chunk below creates a panel data frame using the `pdata.frame()` function from the `plm` package.

```{r}
pdata <- pdata.frame(wide_data, index = c("Region", "Year"))

```

The result of this code is that the `wide_data` data frame is now transformed into a **panel data frame**, stored in the variable `pdata`. This allows we to perform panel data analysis using the `plm` package.

-   The `pdata` object will now be suitable for running panel regressions, such as fixed effects or random effects models, using functions like `plm()` from the `plm` package.

The below code chunk use `colnames()` function to return the column names of the `pdata` which is now a panel data after being converted by `pdata.frame()`.

```{r}
colnames(pdata)

```

The `view()` function will open the `pdata` in a spreadsheet-like viewer in Rstudio.

```{r}
view(pdata)
```

## 4.3 PLM

### 4.3.1 Initial run of the Model to confirm the variables

We will be use the `DiarrheaRate` as a sample to test the performance of our initially selected indicator in the `plm`.

The below code chunk fits a Fixed Effects Model also known as the 'within' model) using the `plm` function from the `plm` package in R.

The **Fixed Effects Model** estimates how changes in the predictor variables (e.g., sanitation services, population density, water stress) influence the **DiarrheaRate** within each `Region` across different years, controlling for unobserved time-invariant characteristics of each region (such as cultural or geographical factors).

```{r}
model <- plm(DiarrheaRate ~ 
  People.practicing.open.defecation....of.population. +
  People.using.at.least.basic.drinking.water.services....of.population. +
  People.using.at.least.basic.sanitation.services....of.population. +
  Population.density..people.per.sq..km.of.land.area. +
  Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal. +
  People.with.basic.handwashing.facilities.including.soap.and.water....of.population. +
  Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources +  # Level of water stress
  People.using.safely.managed.drinking.water.services....of.population. +  # People using safely managed drinking water services
  People.using.safely.managed.sanitation.services....of.population.,  # People using safely managed sanitation services
  data = pdata,
  model = "within"
)
summary(model)

```

Model Details base on the result:

-   Balanced Panel: The dataset used for the model has 2 cross-sectional units (regions) and 22 time periods, resulting in a total of 44 observations.

-   Model: The model used is the Within Model (Fixed Effects), which controls for time-invariant characteristics of each region.

However, we had 6 regions but the result was only 2 regions. It is likely there are missing values in the key variables.

The below code chunk check for the missing data in the `pdata`.

```{r}
colSums(is.na(pdata))

```

The result shows there are 88 missing value in the `People.with.basic.handwashing.facilities.including.soap.and.water....of.population.`

It was decide to remove the `People.with.basic.handwashing.facilities.including.soap.and.water....of.population.` from the model to if there is any improvement.

### 4.3.2 Model run after remove `People.with.basic.handwashing.facilities.including.soap.and.water....of.population.`

The below code chunk fits a Fixed Effects Model also known as the 'within' model) using the `plm` function from the `plm` package in R.

```{r}

model <- plm(DiarrheaRate ~ 
  People.practicing.open.defecation....of.population. +
  People.using.at.least.basic.drinking.water.services....of.population. +
  People.using.at.least.basic.sanitation.services....of.population. +
  Population.density..people.per.sq..km.of.land.area. +
  Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal. +
  Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources +  # Level of water stress
  People.using.safely.managed.drinking.water.services....of.population. +  # People using safely managed drinking water services
  People.using.safely.managed.sanitation.services....of.population.,  # People using safely managed sanitation services
  data = pdata,
  model = "within"
)
summary(model)

```

Model Overview base on the result:

-   Balanced Panel: The dataset consists of 6 regions (cross-sectional units) and 22 time periods, with a total of 132 observations.

-   Model Type: Fixed Effects Model (within), which controls for time-invariant characteristics of each region.

Model Summary:

-   R-Squared: 0.85406, meaning the model explains 85.4% of the variance in DiarrheaRate.

-   Adj. R-Squared: 0.83798, a good adjustment for the number of predictors.

-   F-statistic: 86.3196 (p-value \< 2.22e-16), showing that the model is highly significant.

The fixed effects model indicates that factors like open defecation, basic drinking water services, water stress, and water productivity are significantly related to DiarrheaRate, while factors like population density and safely managed sanitation services are not significant in this context. The high R-squared value suggests that the model is a good fit for the data.\

It was decide to continue to test for removing the non-significant variables. By removing `Population.density` and `People.using.safely.managed.sanitation.services` to simplify the model.

### 4.3.3 Test run removing variables `Population.density` and `People.using.safely.managed.sanitation.services`

The below code chunk fits a Fixed Effects Model also known as the 'within' model) using the `plm` function from the `plm` package in R.

```{r}

model <- plm(DiarrheaRate ~ 
  People.practicing.open.defecation....of.population. +
  People.using.at.least.basic.drinking.water.services....of.population. +
  People.using.at.least.basic.sanitation.services....of.population. +
  Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal. +
  Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources + 
  People.using.safely.managed.drinking.water.services....of.population. ,  
  
  data = pdata,
  model = "within"
)
summary(model)

```

Model Overview:

-   Balanced Panel: The dataset consists of 6 regions (cross-sectional units) and 22 time periods, with a total of 132 observations.

-   Model Type: The model used is the Fixed Effects Model (within), which controls for time-invariant characteristics of each region.

Model Fit:

-   R-Squared: The model explains 85.12% of the variance in DiarrheaRate, indicating a strong fit.

-   Adjusted R-Squared: 83.76%, showing that the model explains most of the variance even after accounting for the number of predictors.

-   F-statistic: 114.42 on 6 and 120 degrees of freedom, with a p-value \< 2.22e-16, confirming that the model is statistically significant.

The fixed effects model indicates that open defecation, drinking water access, and sanitation services are the most significant factors influencing DiarrheaRate across regions. Other factors like water productivity and water stress also have significant effects. The high R-squared value suggests that the model explains most of the variation in diarrhea rates, and the predictors are statistically significant.

### 4.3.4 Model Comparsion

It was then decide to compare the model to decide which will be apply in the further analysis.

The below code chunk fits a fixed Fixed Effects Model using the `plm` function, with the DiarrheaRate as the dependent variables and all the predictors after removing `People.using.safely.managed.sanitation.services`

```{r}
# Original Model (all predictors)
model_full <- plm(DiarrheaRate ~ 
  People.practicing.open.defecation....of.population. +
  People.using.at.least.basic.drinking.water.services....of.population. +
  People.using.at.least.basic.sanitation.services....of.population. +
  Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal. +
  Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources +
  People.using.safely.managed.drinking.water.services....of.population., 
  data = pdata, model = "within")
summary(model_full)

```

The below code chunk fits a fixed Fixed Effects Model using the `plm` function, with the DiarrheaRate as the dependent variables and remove the non-significant variables including `Population.density` and `People.using.safely.managed.sanitation.services` .

```{r}
# Reduced Model (without non-significant predictors)
model_reduced <- plm(DiarrheaRate ~ 
  People.practicing.open.defecation....of.population. +
  People.using.at.least.basic.drinking.water.services....of.population. +
  People.using.at.least.basic.sanitation.services....of.population. +
  Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources +
  People.using.safely.managed.drinking.water.services....of.population., 
  data = pdata, model = "within")
summary(model_reduced)

```

The below code chunk provide the summary of the R-squared, Adjusted R-squared, F-statistic for both models.

```{r}
summary(model_full)$r.squared
summary(model_reduced)$r.squared

summary(model_full)$adj.r.squared
summary(model_reduced)$adj.r.squared

summary(model_full)$fstatistic
summary(model_reduced)$fstatistic

```

The R-Squared value of the reduced model is slightly lower than the full model, which means the full model explains a little more of the variance in DiarrheaRate.

The Adjusted R-Squared also shows a slight drop in the reduced model, though the difference is not large, indicating that removing some predictors did not dramatically reduce the model's explanatory power after adjusting for the number of predictors.

Both models show that at least one of the predictors is statistically significant (given the very low p-values).

The reduced model has a slightly higher F-statistic, which suggests that removing one predictor did not negatively affect the overall model’s significance, and may even have improved it slightly.\
\
The full model explains slightly more variance in DiarrheaRate than the reduced model based on both R-Squared and Adjusted R-Squared.

It was decide we will be continue with our full model.

### 4.3.5 PLM continue with full model

The below code chunk fits a Fixed Effects Model also known as the 'within' model) using the `plm` function from the `plm` package in R analyse the relationship between DiarrheaRate and WASH indicators.

```{r}
model <- plm(DiarrheaRate ~ 
  People.practicing.open.defecation....of.population. +
  People.using.at.least.basic.drinking.water.services....of.population. +
  People.using.at.least.basic.sanitation.services....of.population. +
  Population.density..people.per.sq..km.of.land.area. +
  Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal. +
  Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources +  # Level of water stress
  People.using.safely.managed.drinking.water.services....of.population. +  # People using safely managed drinking water services
  People.using.safely.managed.sanitation.services....of.population.,  # People using safely managed sanitation services
  data = pdata,
  model = "within"
)
summary(model)

```

The below code chunk fits a Fixed Effects Model also known as the 'within' model) using the `plm` function from the `plm` package in R analyse the relationship between `HeptatitisRate` and WASH indicators.

```{r}
model <- plm(HepatitisRate ~ 
  People.practicing.open.defecation....of.population. +
  People.using.at.least.basic.drinking.water.services....of.population. +
  People.using.at.least.basic.sanitation.services....of.population. +
  Population.density..people.per.sq..km.of.land.area. +
  Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal. +
  Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources +  # Level of water stress
  People.using.safely.managed.drinking.water.services....of.population. +  # People using safely managed drinking water services
  People.using.safely.managed.sanitation.services....of.population.,  # People using safely managed sanitation services
  data = pdata,
  model = "within"
)
summary(model)

```

The below code chunk fits a Fixed Effects Model also known as the 'within' model) using the `plm` function from the `plm` package in R analyse the relationship between `unsafetyRisk` and WASH indicators.

```{r}
model <- plm(UnsafeRisk ~ 
  People.practicing.open.defecation....of.population. +
  People.using.at.least.basic.drinking.water.services....of.population. +
  People.using.at.least.basic.sanitation.services....of.population. +
  Population.density..people.per.sq..km.of.land.area. +
  Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal. +
  Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources +  # Level of water stress
  People.using.safely.managed.drinking.water.services....of.population. +  # People using safely managed drinking water services
  People.using.safely.managed.sanitation.services....of.population.,  # People using safely managed sanitation services
  data = pdata,
  model = "within"
)
summary(model)
```

The below code chunk fits a Fixed Effects Model also known as the 'within' model) using the `plm` function from the `plm` package in R analyse the relationship between `TyphoidRate` and WASH indicators.

```{r}
model <- plm(TyphoidRate~ 
  People.practicing.open.defecation....of.population. +
  People.using.at.least.basic.drinking.water.services....of.population. +
  People.using.at.least.basic.sanitation.services....of.population. +
  Population.density..people.per.sq..km.of.land.area. +
  Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal. +
  Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources +  # Level of water stress
  People.using.safely.managed.drinking.water.services....of.population. +  # People using safely managed drinking water services
  People.using.safely.managed.sanitation.services....of.population.,  # People using safely managed sanitation services
  data = pdata,
  model = "within"
)
summary(model)


```

Base on the output of four model, People practicing open defecation and People using at least basic drinking water services are significant across all models, consistently showing a negative relationship with health outcomes. `Level of water stress` appears as significant in most models, suggesting that regions facing higher water stress tend to experience worse health outcomes. `People using safely managed drinking water services` has significant positive impacts on `DiarrheaRate` and `UnsafeRisk`, but it is not significant for `HepatitisRate`.

The models explaining `HepatitisRate` and `UnsafeRisk` have very high R-squared values (above 94%), suggesting that these models explain most of the variance in the respective dependent variables.The `DiarrheaRate` model has a slightly lower R-squared (85.4%), but still a very good fit.

The results show that sanitation, drinking water access, and open defecation are important factors influencing health outcomes, particularly `DiarrheaRate` and `UnsafeRisk`.Water stress plays a critical role, indicating that regions with high water stress are likely to face more health-related challenges.There is considerable variation in the impact of predictors on different health outcomes, with some predictors (e.g., `People using safely managed drinking water services`) being more influential for some health outcomes than others.

## 4.4 Coefficient list

The code chunk below create a list of coefficient for four different health outcome, `DiarrheaRate`, `TyphoidRate`, `UnsafeRisk`, and `HepatitisRate` from the fixed effects models.

```{r}
coef_list <- list(
  DiarrheaRate = c(
    OpenDefecation = -0.6755407,
    BasicWater = -1.0144098,
    BasicSanitation = -0.1706479,
    WaterProductivity = 0.0029523,
    WaterStress = 0.0110199,
    PopulationDensity = 0.0014148,
    SafelyManagedWater = 0.2855646,
    SafelyManagedSanitation = -0.1302516
  ),
  TyphoidRate = c(
    OpenDefecation = 0.015961,
    BasicWater = 0.018088,
    BasicSanitation = 0.0024683,
    WaterProductivity = 1.1884e-05,
    WaterStress = -4.1832e-04,
    PopulationDensity = 9.1629e-05,
    SafelyManagedWater = -4.9622e-03,
    SafelyManagedSanitation = -2.0081e-03
  ),
  UnsafeRisk = c(
    OpenDefecation = 0.72708968,
    BasicWater = 0.43477443,
    BasicSanitation = 0.21651056,
    WaterProductivity = 0.00063762,
    WaterStress = -0.01913748,
    PopulationDensity = 0.00037018,
    SafelyManagedWater = -0.25010937,
    SafelyManagedSanitation = 0.04852111
  ),
  HepatitisRate = c(
    OpenDefecation = 0.0085691,
    BasicWater = 0.023667,
    BasicSanitation = -0.011585,
    WaterProductivity = 8.6247e-05,
    WaterStress = -0.0010463,
    PopulationDensity = -3.7942e-04,
    SafelyManagedWater = -3.6054e-03,
    SafelyManagedSanitation = -5.7112e-04
  )
)

```

# 5. Time Series Analysis for period 2022 to 2030 for wash indicators

## 5.1 Create `selected_data`

The below code chunk use `dplyr` package to select all the variables that will be use for analysis in the `pdata`. This is done using the `select()` function.

```{r}
selected_data <- pdata %>%
  select(Region, Year, TyphoidRate, DiarrheaRate, HepatitisRate, 
         UnsafeRisk, 
         "Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources",
         "People.practicing.open.defecation....of.population.",
         "People.using.at.least.basic.drinking.water.services....of.population.",
         "People.using.at.least.basic.sanitation.services....of.population.",
         "People.using.safely.managed.drinking.water.services....of.population.",
         "People.using.safely.managed.sanitation.services....of.population.",
         "Population.density..people.per.sq..km.of.land.area.",
         "Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal.")

```

The `view(selected_data)` function in R will open a spreadsheet-like view of the `selected_data` data frame in RStudio, allowing interactively explore and inspect the data.

```{r}
view(selected_data)
```

### 5.2 Converting Data type

The function `class()` and `typeof()` provide different types of information about the Year column in the `selected_data.`

```{r}
class(selected_data$Year)
typeof(selected_data$Year)
```

The below code chunk ensure that if the `Year` column in the `selected_data` is a factor it will be converted into a numeric format.

```{r}
selected_data$Year <- as.numeric(as.character(selected_data$Year))
```

The below code chunk use function `extract_year()` that extracts a 4-digit year from a mixed string and then cleans the `Year` column in your `selected_data` dataset by applying this function. Finally, it converts the `Year` to a Date format representing the first day of the year.

```{r}
extract_year <- function(x) {
  year <- str_extract(x, "\\d{4}")
  return(as.numeric(year))
}

selected_data$Year <- extract_year(selected_data$Year)

selected_data$YearDate <- as.Date(paste0(selected_data$Year, "-01-01"))
```

The below code chunk using str() function provides a compact display the structure of the an object, including its type and other relevant details. The `str(selected_data$Year)`

```{r}
# Check the type of the 'Year' column
str(selected_data$Year)

```

### 5.3 Times Series Analysis using ARIMA

The below code chunk perform ARIMA (Auto-Regression Integrated Moving Average) time series forecasting for WHO regions based on historical data. It generate the WASH indicators and population from 2022 to 2030.\
The code chunk processes data region-by-region, checks the availability of necessary data, fits ARIMA models, and returns the forecasted values for the next several years, which can be used for policy planning, resource allocation, or trend analysis.

```{r}
arima_forecast <- function(data, forecast_start_year = 2022, forecast_end_year = 2030) {

#define variable name 
  forecast_vars <- c(
    "WaterStress" = "Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources",
    "OpenDefecation" = "People.practicing.open.defecation....of.population.",
    "BasicWaterServices" = "People.using.at.least.basic.drinking.water.services....of.population.",
    "BasicSanitationServices" = "People.using.at.least.basic.sanitation.services....of.population.",
    "SafeManagedWaterServices" = "People.using.safely.managed.drinking.water.services....of.population.",
    "SafeManagedSanitationServices" = "People.using.safely.managed.sanitation.services....of.population.",
    "PopulationDensity" = "Population.density..people.per.sq..km.of.land.area.",
    "WaterProductivity" = "Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal."
  )
  
  # Print progress header
  cat("========================================\n")
  cat("Starting ARIMA forecasting for", forecast_start_year, "to", forecast_end_year, "\n")
  cat("========================================\n\n")
  
  # Verify input data
  cat("Verifying input data...\n")
  cat("Input data has", nrow(data), "rows and", ncol(data), "columns\n")
  
  # Check if required columns exist
  required_cols <- c("Region", "Year")
  missing_required <- setdiff(required_cols, names(data))
  if (length(missing_required) > 0) {
    stop("Missing required columns: ", paste(missing_required, collapse = ", "))
  }
  
  # Check for variable columns
  available_vars <- intersect(forecast_vars, names(data))
  if (length(available_vars) == 0) {
    stop("None of the forecast variables found in the data")
  }
  cat("Found", length(available_vars), "out of", length(forecast_vars), "forecast variables\n")
  
  # Get unique regions
  regions <- unique(data$Region)
  cat("Found", length(regions), "regions:", paste(regions, collapse = ", "), "\n\n")
  
  # Create empty data frame for results
  all_forecasts <- data.frame()
  
  # Process each region
  for (region in regions) {
    cat("Processing region:", region, "\n")
    
    # Filter data for this region
    region_data <- data %>% filter(Region == region)
    cat("  Region has", nrow(region_data), "rows\n")
    
    # Create an empty data frame for this region's forecasts
    region_forecasts <- expand.grid(
      Region = region,
      Year = forecast_start_year:forecast_end_year,
      stringsAsFactors = FALSE
    )
    # Add DataType column
    region_forecasts$DataType <- "Forecast"
    
    # Process each variable
    for (short_name in names(forecast_vars)) {
      var_name <- forecast_vars[short_name]
      cat("  Forecasting:", short_name, "(", var_name, ")\n")
      
      # Check if variable exists in data
      if (!(var_name %in% names(region_data))) {
        cat("    Variable not found in data, skipping\n")
        region_forecasts[[short_name]] <- NA
        next
      }
      
      # Get historical data for this variable
      var_data <- region_data %>% 
        filter(!is.na(!!sym(var_name)), Year < forecast_start_year) %>%
        arrange(Year)
      
      # Check if enough data
      if (nrow(var_data) < 5) {
        cat("    Insufficient historical data (need at least 5 points), skipping\n")
        region_forecasts[[short_name]] <- NA
        next
      }
      
      # Create time series
      cat("    Creating time series with", nrow(var_data), "points from", 
          min(var_data$Year), "to", max(var_data$Year), "\n")
      
      ts_data <- ts(var_data[[var_name]], 
                   start = min(var_data$Year), 
                   end = max(var_data$Year), 
                   frequency = 1)
      
      # Fit ARIMA model and forecast
      tryCatch({
        cat("    Fitting ARIMA model...\n")
        arima_model <- auto.arima(ts_data, seasonal = FALSE, stepwise = TRUE)
        
        # Print model information
        cat("    Model:", arimaorder(arima_model), "\n")
        
        # Generate forecast
        forecast_years <- forecast_end_year - forecast_start_year + 1
        cat("    Generating", forecast_years, "year forecast...\n")
        forecast_result <- forecast(arima_model, h = forecast_years)
        
        # Add forecast to results
        region_forecasts[[short_name]] <- as.numeric(forecast_result$mean)
        
        # Print forecast summary
        cat("    Forecast summary - First:", round(forecast_result$mean[1], 4),
            "Last:", round(forecast_result$mean[forecast_years], 4), "\n")
        
      }, error = function(e) {
        cat("    Error in forecasting:", e$message, "\n")
        region_forecasts[[short_name]] <- NA
      })
    }
    
    # Add this region's forecasts to the overall results
    all_forecasts <- bind_rows(all_forecasts, region_forecasts)
    cat("  Completed forecasts for", region, "\n\n")
  }
  
  # Print summary
  cat("========================================\n")
  cat("ARIMA forecasting complete\n")
  cat("Generated forecasts for", nrow(all_forecasts), "rows\n")
  cat("Forecasted variables:", paste(names(all_forecasts)[!(names(all_forecasts) %in% c("Region", "Year", "DataType"))], collapse = ", "), "\n")
  cat("========================================\n\n")
  
  # Return the clean format results with short names
  return(all_forecasts)
}


```

The below code chunk run the `arima_forecast` from above code chunk, generating forecasts for year 2022 to 2030. The resulting forecasted data is then saved to a csv file name `water_sanitation_forecasts.csv` .

```{r}

 forecast_data <-arima_forecast(selected_data, 
                                         forecast_start_year = 2022, 
                                          forecast_end_year = 2030)
 
 write.csv(forecast_data, "water_sanitation_forecasts.csv", row.names = FALSE)
```

The `view(forecast_data)` command will open a viewer in RStudio, displaying the `forecast_data` dataset in a spreadsheet-like format.

```{r}
view(forecast_data)
```

# 6. Forecast for health indicator

## 6.1 Apply coefficient to forecast data

The below code chunk generates predictions for disease rates (Diarrhea, Typhoid, Unsafe Risk, and Hepatitis) based on forecasted changes in key water and sanitation variables (e.g., open defecation, basic water services, population density) for each region. The function first retrieves baseline values from the historical data for 2021, then calculates the forecasted disease rates for each region between 2022 and 2030 by applying coefficients that quantify the impact of changes in the variables. It iterates over each region and variable, adjusting the baseline values according to the forecasted changes, and ensures that negative disease values are set to zero. The function returns the updated dataset with forecasted disease rates.

```{r}
predict_diseases <- function(forecast_data, selected_data) {
  var_mapping <- c(
    "Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources" = "WaterStress",
    "People.practicing.open.defecation....of.population." = "OpenDefecation",
    "People.using.at.least.basic.drinking.water.services....of.population." = "BasicWaterServices",
    "People.using.at.least.basic.sanitation.services....of.population." = "BasicSanitationServices",
    "People.using.safely.managed.drinking.water.services....of.population." = "SafeManagedWaterServices",
    "People.using.safely.managed.sanitation.services....of.population." = "SafeManagedSanitationServices",
    "Population.density..people.per.sq..km.of.land.area." = "PopulationDensity",
    "Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal." = "WaterProductivity"
  )
  

  historical <- as.data.frame(selected_data)
  

  baseline <- historical[historical$Year == 2021, ]
  

  coef_list <- list(
    DiarrheaRate = c(
      OpenDefecation = -0.6755407,
      BasicWaterServices = -1.0144098,
      BasicSanitationServices = -0.1706479,
      WaterProductivity = 0.0029523,
      WaterStress = 0.0110199,
      PopulationDensity = 0.0014148,
      SafeManagedWaterServices = 0.2855646,
      SafeManagedSanitationServices = -0.1302516
    ),
    TyphoidRate = c(
      OpenDefecation = 0.015961,
      BasicWaterServices = 0.018088,
      BasicSanitationServices = 0.0024683,
      WaterProductivity = 1.1884e-05,
      WaterStress = -4.1832e-04,
      PopulationDensity = 9.1629e-05,
      SafeManagedWaterServices = -4.9622e-03,
      SafeManagedSanitationServices = -2.0081e-03
    ),
    UnsafeRisk = c(
      OpenDefecation = 0.72708968,
      BasicWaterServices = 0.43477443,
      BasicSanitationServices = 0.21651056,
      WaterProductivity = 0.00063762,
      WaterStress = -0.01913748,
      PopulationDensity = 0.00037018,
      SafeManagedWaterServices = -0.25010937,
      SafeManagedSanitationServices = 0.04852111
    ),
    HepatitisRate = c(
      OpenDefecation = 0.0085691,
      BasicWaterServices = 0.023667,
      BasicSanitationServices = -0.011585,
      WaterProductivity = 8.6247e-05,
      WaterStress = -0.0010463,
      PopulationDensity = -3.7942e-04,
      SafeManagedWaterServices = -3.6054e-03,
      SafeManagedSanitationServices = -5.7112e-04
    )
  )
  
  
  result <- forecast_data
  
 
  for (disease in names(coef_list)) {
    result[[disease]] <- NA
  }
  

  for (region in unique(result$Region)) {

    region_baseline <- baseline[baseline$Region == region, ]
    
    if (nrow(region_baseline) == 0) {
      next
    }
    
 
    region_rows <- which(result$Region == region)
    
   
    for (disease in names(coef_list)) {
   
      baseline_value <- region_baseline[[disease]]
      
      for (row in region_rows) {
      
        result[row, disease] <- baseline_value
        
        for (pred_name in names(coef_list[[disease]])) {
          
          forecast_col <- pred_name
          
          historical_col <- names(var_mapping)[var_mapping == forecast_col]
          
          if (length(historical_col) == 0 || !forecast_col %in% names(result)) {
            next
          }
          
          baseline_val <- region_baseline[[historical_col]]
          forecast_val <- result[row, forecast_col]
          
          if (is.na(baseline_val) || is.na(forecast_val)) {
            next
          }
          
          coef_val <- coef_list[[disease]][pred_name]
          change <- forecast_val - baseline_val
          effect <- coef_val * change
          result[row, disease] <- result[row, disease] + effect
        }
      }
    }
  }
  
  # Set all negative disease values to zero
  for (disease in names(coef_list)) {
    negative_rows <- which(result[[disease]] < 0)
    if (length(negative_rows) > 0) {
      result[negative_rows, disease] <- 0
    }
  }
  
  return(result)
}
```

The code chunk below runs the `predict_diseases` created above, generating the predicted disease rates for each region and year. It then saves the results to a csv file named `disease_predictions.csv`.

```{r}
disease_forecasts <- predict_diseases(forecast_data, selected_data)


write.csv(disease_forecasts, "disease_predictions.csv", row.names = FALSE)
```

The `view(disease_forecasts)` function will open a spreadsheet-like viewer in RStudio, displaying the contents of the **`disease_forecasts`** data frame.

```{r}
view(disease_forecasts)
```

# 7.Merging Historical data with Forcasted data

The code combines historical and forecasted data by first mapping the original water and sanitation variables from the `selected_data` dataset to standardized names, then cleaning and adding relevant columns for disease rates (e.g., DiarrheaRate, TyphoidRate) and water-related variables (e.g., WaterStress, PopulationDensity). It adds a `DataType` column to distinguish historical from forecast data, merges both datasets, and sorts them by Region and Year. Finally, the combined data is saved as a CSV file, `combined_complete_data.csv`, containing both historical and forecasted values for further analysis or reporting.

```{r}

historical_vars <- c(
  "Level.of.water.stress..freshwater.withdrawal.as.a.proportion.of.available.freshwater.resources",
  "People.practicing.open.defecation....of.population.",
  "People.using.at.least.basic.drinking.water.services....of.population.",
  "People.using.at.least.basic.sanitation.services....of.population.",
  "People.using.safely.managed.drinking.water.services....of.population.",
  "People.using.safely.managed.sanitation.services....of.population.",
  "Population.density..people.per.sq..km.of.land.area.",
  "Water.productivity..total..constant.2015.US..GDP.per.cubic.meter.of.total.freshwater.withdrawal."
)

forecast_vars <- c(
  "WaterStress",
  "OpenDefecation",
  "BasicWaterServices",
  "BasicSanitationServices",
  "SafeManagedWaterServices", 
  "SafeManagedSanitationServices",
  "PopulationDensity",
  "WaterProductivity"
)


var_mapping <- setNames(forecast_vars, historical_vars)


historical_data <- as.data.frame(selected_data)

historical_data$DataType <- "Historical"


historical_clean <- historical_data[, c("Region", "Year", "TyphoidRate", "DiarrheaRate", 
                                       "HepatitisRate", "UnsafeRisk", "DataType")]


for (hist_var in names(var_mapping)) {
  if (hist_var %in% names(historical_data)) {
    forecast_var <- var_mapping[hist_var]
    historical_clean[[forecast_var]] <- historical_data[[hist_var]]
  }
}


combined_data <- rbind(historical_clean, disease_forecasts)


combined_data <- combined_data[order(combined_data$Region, combined_data$Year), ]

```

```{r}
write.csv(combined_data, "combined_complete_data.csv", row.names = FALSE)
```

# 8. Visualize test result

## 8.1 Importing the combined data

The below code chunk use `read.csv()` to read the `combined_complete_data.csv`.

```{r}
combined_data <- read.csv("combined_complete_data.csv")
```

## 8.2 plotting for visualization

The below code chunk creates a line plot using `ggplot2` to visualise the historical result and forecast result of the `DiarrheaRate`

```{r}
ggplot(combined_data, aes(x = Year, y = DiarrheaRate, color = Region, linetype = DataType)) +
  geom_line() +
  geom_point() +
  labs(title = "Diarrhea Rate: Historical and Forecast", y = "Rate")
```

```{r}

combined_data_long <- combined_data %>%
  pivot_longer(cols = c(UnsafeRisk, WaterStress, OpenDefecation, BasicWaterServices, 
                        BasicSanitationServices, SafeManagedWaterServices, 
                        SafeManagedSanitationServices, WaterProductivity),
               names_to = "WASH_Variable", values_to = "Value")
```

```{r}

ggplot(combined_data_long, aes(x = Year, y = Value, color = Region, linetype = DataType)) +
  geom_line() +
  geom_point() +
  facet_wrap(~WASH_Variable, scales = "free_y") +  # Facet by WASH variable
  labs(title = "WASH Variables Over Time by Region", y = "Value") +
  theme_minimal()


```

```{r}
library(ggplot2)
library(tidyr)

# Load the combined data
combined_data <- read.csv("combined_complete_data.csv")

# Reshape the data to a longer format for both WASH and disease variables
combined_data_long <- combined_data %>%
  pivot_longer(cols = c(UnsafeRisk, WaterStress, OpenDefecation, BasicWaterServices, 
                        BasicSanitationServices, SafeManagedWaterServices, 
                        SafeManagedSanitationServices, WaterProductivity,
                        TyphoidRate, DiarrheaRate, HepatitisRate),
               names_to = "Variable", values_to = "Value")

# Plot all WASH and disease variables for all regions
ggplot(combined_data_long, aes(x = Year, y = Value, color = Region, linetype = DataType)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Variable, scales = "free_y") +  # Facet by variable (WASH + diseases)
  labs(title = "WASH and Disease Variables Over Time by Region", y = "Value") +
  theme_minimal()

```
