---
title: "Take Home Exercise 3"
author: "Lai Meihao"
date: "March 22, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

1.  Overview

1.1 Project Brief

Take home exercise 3 is a preliminary work of the group project - blue pulse. Water is essential to life, yet billions worldwide still lack access to safe drinking water and sanitation. Waterborne disease such as acute hepatitis A and diarrhea continue to pose series public health threats, particularly in regions with inadequate water access. SDG 6 aims to ensure universal access to water and sanitation The project aims to understand the 1) understand the distribution of water access and safety by region, 2) analyze the long-term effects of water access on health outcome.

The project team consists of three members and each member will take one of the main prototype modules as follows:

-   Exploratory Data Analysis (initial & treemap)

-   Panel Linear Model for observations over time for multiple entities (regions, countries)

-   Confirmatory Data Analysis.

1.2 Project Objectives

The project will using WDI from world bank and burden of disease from institute for health metrics and evaluation. The objective of this take-home exercise is to build a prototype - user interface design on Panel Linear Model.

```{r}
pacman::p_load(tidyverse, ggplot2, plm,readxl)
```

```{r}
WDI <- read.csv("data/cd584ca9-08f7-46dd-87c1-00ce4af84b71_Data.csv")
```

```{r}
Disease <- "data/WaterandDiseasebywhoregion.xlsx"

sheet_names <- excel_sheets(Disease)

data_excel <- lapply(sheet_names, function(sheet) {
  read_excel(Disease, sheet = sheet)
})

names(data_excel) <- sheet_names

print(names(data_excel)) 
str(data_excel)          

```
