---
title: "Take-home Exercise 3 Part 2"
author: "Lai Meihao"
date: "March 27, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1.Overview

This part focus on creating visualization for the `combined_complete_data.csv`.

# 2.Data Preparation

## 2.1 Install and launch R packages

The project uses `p_load()` of pacman package to check if the R package are installed in the computer.

The following code chunk is used to install and launch the R packages.

```{r}
pacman::p_load(tidyverse, ggplot2, plm,ExPanDaR,lubridate, ggthemes, plotly,purrr,shiny)
```

## 2.2 Import Data

The below code chunk use `read_csv()` to read the csv file into R as a data frame

```{r}
data <- read_csv("combined_complete_data.csv")
```

# 3.Plotting Scatter Plot for Africa Region

The below code chunk plot a scatter for Africa between `BasicSanitationService` and `DiarrheaRate`

```{r}

africa_data <- data %>%
  filter(Region == "African Region")

africa_scatter <- ggplot(africa_data, 
                         aes(x = BasicSanitationServices, 
                             y = DiarrheaRate, 
                             color = DataType)) +
  
  geom_point(aes(size = PopulationDensity), 
             alpha = 0.7) +
  
  
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "red", 
              linetype = "dashed") +
  

  labs(
    title = "Diarrhea Rate vs Basic Sanitation Services in African Region",
    x = "Basic Sanitation Services (% of population)",
    y = "Diarrhea Rate (%)",
    color = "Data Type"
  ) +
  

  theme_minimal() +
  
  scale_color_brewer(palette = "Set1")


print(africa_scatter)


regression_model <- lm(DiarrheaRate ~ BasicSanitationServices, data = africa_data)
print("Regression Summary:")
summary(regression_model)


print("Correlation Details:")
cor_results <- cor.test(africa_data$BasicSanitationServices, 
                        africa_data$DiarrheaRate)
print(cor_results)
```

# 4 Use ExPanD to assist in determine the visualisation for shiny

## 4.1 Overview of the ExPanD package

The `ExPanDaR` package is designed for exploratory panel data analysis in R. It provides a set of tools that allow users to conduct various analyses on panel data, which involves data collected across time and across different subjects (such as countries, individuals, etc.).

The package aims to simplify and streamline tasks like:

1.  **Data exploration**: Summarizing panel data, understanding the structure, and generating basic descriptive statistics.

2.  **Visualization**: Creating plots and charts to visualize trends, distributions, and relationships in panel data.

3.  **Modeling**: Performing regression analyses on panel data, including fixed-effects and random-effects models.

4.  **Diagnostics**: Conducting diagnostic checks to assess the assumptions of panel data models.

## 4.2 Using `ExPanD()`

The below code chunk use the `ExpanD()` function, and set `cs_id="Region"`, the `ts_id="Year"`. The cs_id refers to the cross-sectional identifier (the entity being observed across time). In this case, it’s set to "Region", indicating that each region represents an individual subject (or entity) observed over multiple time periods.The `ts_id` refers to the time-series identifier, specifying that the "Year" column contains the time periods in which the data was collected.

```{r}
#remove the '#' for the line below for using ExPanD
#ExPanD(df = data, cs_id = "Region", ts_id = "Year")
```

After review the output, it was decide to proceed with time trend and bar chart with multiple selection for variables.

# 5. Testing plots for visualization

## 5.1 plot with ggplot2

```{r}
data$Year <- as.numeric(data$Year)
data$DataSource <- ifelse(data$Year > 2020, "Forecast", "Actual")


filtered_data <- function(variableType, washVarsSelect, diseaseSelect, regionSelect, dataSourceSelect, yearSelect) {
  data_filtered <- data
  

  data_filtered <- data_filtered %>%
    filter(Year >= yearSelect[1] & Year <= yearSelect[2]) %>%
    filter(Region %in% regionSelect)
 
  data_filtered <- data_filtered %>%
    filter(DataSource %in% dataSourceSelect)
  
  selected_vars <- if (variableType == "WASH Variables") {
    washVarsSelect
  } else if (variableType == "Disease Rates") {
    diseaseSelect
  } else {  # Compare Both
    c(washVarsSelect, diseaseSelect)
  }
  
  plot_data <- data_filtered %>%
    select(Region, Year, DataSource, all_of(selected_vars)) %>%
    pivot_longer(
      cols = all_of(selected_vars),
      names_to = "Variable",
      values_to = "Value"
    )
  
  return(plot_data)
}


```

```{r}
#line plot
line_plot <- function(plot_data) {
  ggplot(plot_data, aes(x = Year, y = Value, color = Variable)) +
    geom_line(size = 1) +
    labs(title = "Line Plot of Selected Variables", x = "Year", y = "Value") +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

```{r}
#box plot
box_plot <- function(plot_data) {
  ggplot(plot_data, aes(x = Variable, y = Value, fill = Variable)) +
    geom_boxplot() +
    labs(title = "Box Plot of Selected Variables", x = "Variable", y = "Value") +
    theme_minimal()
}
```

```{r}
#histogram
histogram_plot <- function(plot_data) {
  ggplot(plot_data, aes(x = Value, fill = Variable)) +
    geom_histogram(binwidth = 0.1, alpha = 0.6, position = "identity") +
    labs(title = "Histogram of Selected Variables", x = "Value", y = "Frequency") +
    theme_minimal()
}

```

```{r}
#facet grid
facet_plot <- function(plot_data) {
  ggplot(plot_data, aes(x = Year, y = Value, color = Variable)) +
    geom_line(size = 1) +
    facet_wrap(~Region) +  # Facet by region or other categories
    labs(title = "Facet Grid of Selected Variables", x = "Year", y = "Value") +
    theme_minimal()
}

```

```{r}
# Time Series Plot
time_series_plot <- function(plot_data) {
  ggplot(plot_data, aes(x = Year, y = Value, color = Variable, linetype = DataSource)) +
    geom_line(size = 1, alpha = 0.6) +
    geom_point(size = 2) +
    aes(color = interaction(Variable, Region)) + 
    labs(title = "Time Series of Selected Variables", y = "Value") +
    theme_minimal() +
    theme(legend.position = "bottom") +
    scale_linetype_manual(values = c("solid", "dashed"))
}

```

```{r}
# Scatter Plot
scatter_plot <- function(plot_data) {
  ggplot(plot_data, aes(x = Year, y = Value, color = Variable, shape = DataSource)) +
    geom_point(size = 3) +
    labs(title = "Scatter Plot of Selected Variables", x = "Year", y = "Value") +
    theme_minimal()
}

```

```{r}
# Heatmap
heatmap_plot <- function(plot_data) {
  ggplot(plot_data, aes(x = Year, y = Variable, fill = Value)) +
    geom_tile() +
    labs(title = "Heatmap of Selected Variables", x = "Year", y = "Variable") +
    scale_fill_viridis_c() +
    theme_minimal()
}

```

```{r}
# Bar Chart
bar_chart <- function(plot_data) {
  ggplot(plot_data, aes(x = Year, y = Value, fill = Variable)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = "Bar Chart of Selected Variables", x = "Year", y = "Value") +
    theme_minimal()
}

```

```{r}
density_plot <- function(plot_data) {
  ggplot(plot_data, aes(x = Value, fill = Variable)) +
    geom_density(alpha = 0.6) +
    labs(title = "Density Plot of Selected Variables", x = "Value", y = "Density") +
    theme_minimal()
}

```

```{r}
generate_plots <- function(variableType, washVarsSelect, diseaseSelect, regionSelect, dataSourceSelect, yearSelect) {
  plot_data <- filtered_data(variableType, washVarsSelect, diseaseSelect, regionSelect, dataSourceSelect, yearSelect)
  
  # Time Series Plot
  print(time_series_plot(plot_data))
  
  # Scatter Plot
  print(scatter_plot(plot_data))
  
  # Bar Chart
  print(bar_chart(plot_data))
  
  # Heatmap
  print(heatmap_plot(plot_data))
  
  #box plot
  print(box_plot(plot_data))
  
  #histogram
  print(histogram_plot(plot_data))
  
  #facet_plot
  print(facet_plot(plot_data))
  
  #density plot
  print(density_plot(plot_data))

}
```

```{r}
generate_plots(
  variableType = "Compare Both", 
  washVarsSelect = c("BasicWaterServices", "WaterStress","OpenDefecation"), 
  diseaseSelect = c("TyphoidRate", "DiarrheaRate"), 
  regionSelect = c("African Region"), 
  dataSourceSelect = c("Actual", "Forecast"), 
  yearSelect = c(2000, 2030)
)
```

## 5.2 Interactive Plot

```{r}

interactive_plot <- plot_ly(data, 
                            x = ~Year, 
                            y = ~TyphoidRate, 
                            color = ~Region,   
                            type = 'scatter', 
                            mode = 'lines+markers',  
                            hoverinfo = 'text', 
                            text = ~paste('Year: ', Year, '<br>Typhoid Rate: ', round(TyphoidRate, 2))) %>%
  layout(title = "Interactive Time Series of Typhoid Rate by Region",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Typhoid Rate"))


interactive_plot


```

```{r}

interactive_plot <- plot_ly(data, 
                            x = ~Year, 
                            y = ~TyphoidRate, 
                            color = ~Region,   
                            type = 'scatter', 
                            mode = 'lines+markers', 
                            hoverinfo = 'text', 
                            text = ~paste('Year: ', Year, '<br>Typhoid Rate: ', round(TyphoidRate, 2))) %>%
  layout(title = "Interactive Time Series of Typhoid Rate by Region",
         xaxis = list(title = "Year",
                      rangeslider = list(type = "date"),  # Add range slider
                      type = "linear"),  # Linear scale for year axis
         yaxis = list(title = "Typhoid Rate"),
         dragmode = "zoom")  
interactive_plot

```

```{r}

wash_indicator <- "WaterStress"

interactive_plot <- plot_ly(data, 
                            x = ~Year, 
                            y = ~get(wash_indicator),  
                            color = ~Region,   
                            type = 'scatter', 
                            mode = 'lines+markers',  
                            hoverinfo = 'text', 
                            text = ~paste('Year: ', Year, '<br>', wash_indicator, ': ', round(get(wash_indicator), 2))) %>%
  layout(title = paste("Interactive Time Series of", wash_indicator, "by Region"),
         xaxis = list(title = "Year",
                      rangeslider = list(type = "date"),  
                      type = "linear"),  
         yaxis = list(title = wash_indicator),
         dragmode = "pan")

interactive_plot

```

# 6.Shiny

```{r}
library(shiny)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(tidyr)


combined_df <- read.csv("combined_complete_data.csv", stringsAsFactors = FALSE)

# Ensure Year is numeric and add a DataSource column if not already present
combined_df$Year <- as.numeric(combined_df$Year)
combined_df$DataSource <- ifelse(combined_df$Year > 2020, "Forecast", "Actual")

# Define the UI
ui <- fluidPage(
  titlePanel("WASH Variables & Disease Rates Visualization"),
  
  sidebarLayout(
    sidebarPanel(
      # Variable type selection
      radioButtons("variableType", 
                   "Select Variable Type:",
                   choices = c("WASH Variables", "Disease Rates", "Compare Both"),
                   selected = "Compare Both"),
      
      # WASH variable selection
      conditionalPanel(
        condition = "input.variableType == 'WASH Variables' || input.variableType == 'Compare Both'",
        selectInput("washVarsSelect", 
                    "Select WASH Variables:",
                    choices = c("WaterStress", "OpenDefecation", "BasicWaterServices", 
                                "BasicSanitationServices", "SafeManagedWaterServices", 
                                "SafeManagedSanitationServices", "PopulationDensity", 
                                "WaterProductivity"),
                    selected = c("BasicWaterServices", "WaterStress"),
                    multiple = TRUE)
      ),
      
      # Disease selection
      conditionalPanel(
        condition = "input.variableType == 'Disease Rates' || input.variableType == 'Compare Both'",
        selectInput("diseaseSelect", 
                    "Select Disease Rates:",
                    choices = c("TyphoidRate", "DiarrheaRate", "HepatitisRate", "UnsafeRisk"),
                    selected = "TyphoidRate",
                    multiple = TRUE)
      ),
      
      # Region selection
      selectInput("regionSelect", 
                  "Select Regions:",
                  choices = unique(combined_df$Region),
                  selected = "African Region",
                  multiple = TRUE),
      
      # Visualization type
      selectInput("visualizationType", 
                  "Visualization Type:",
                  choices = c("Time Series", "Scatter Plot", "Bar Chart", "Heatmap"),
                  selected = "Time Series"),
      
      # Data source selection
      checkboxGroupInput("dataSourceSelect",
                         "Data Source:",
                         choices = c("Actual", "Forecast"),
                         selected = c("Actual", "Forecast")),
      
      # Time selection
      sliderInput("yearSelect",
                  "Select Years:",
                  min = min(combined_df$Year),
                  max = max(combined_df$Year),
                  value = c(min(combined_df$Year), max(combined_df$Year)),
                  step = 1, ticks = TRUE)
    ),
    
    mainPanel(
      tabsetPanel(
        tabPanel("Visualization", 
                 plotlyOutput("mainPlot", height = "600px")),
        
        tabPanel("Correlation Analysis",
                 plotlyOutput("correlationPlot", height = "600px")),
        
        tabPanel("Data Table", 
                 DT::dataTableOutput("dataTable"))
      )
    )
  )
)

# Define server function
server <- function(input, output, session) {
  
  # Reactive expression for filtered data
  filtered_data <- reactive({
    data <- combined_df
    # Apply year filter
    data <- data %>%
      filter(Year >= input$yearSelect[1] & Year <= input$yearSelect[2]) %>%
      filter(Region %in% input$regionSelect)
    
    # Filter by data source (Actual/Forecast)
    data <- data %>%
      filter(DataSource %in% input$dataSourceSelect)
    
    return(data)
  })
  
  # Main plot rendering
  output$mainPlot <- renderPlotly({
    data <- filtered_data()
    
    # Check if data is available
    if (nrow(data) == 0) {
      return(ggplotly(ggplot() + 
                      annotate("text", x = 0.5, y = 0.5, label = "No data to display") + 
                      theme_void()))
    }
    
    selected_vars <- if (input$variableType == "WASH Variables") {
      input$washVarsSelect
    } else if (input$variableType == "Disease Rates") {
      input$diseaseSelect
    } else {  # Compare Both
      c(input$washVarsSelect, input$diseaseSelect)
    }
    
    # Reshape data for plotting if multiple variables are selected
    plot_data <- data %>%
      select(Region, Year, DataSource, all_of(selected_vars)) %>%
      pivot_longer(
        cols = all_of(selected_vars),
        names_to = "Variable",
        values_to = "Value"
      )
    
    # Choose plot based on selected visualization type
    if (input$visualizationType == "Time Series") {
      p <- ggplot(plot_data, aes(x = Year, y = Value, color = Variable, linetype = DataSource)) +
        geom_line(size = 1, alpha = 0.6) +  # Make lines less bright
        geom_point(size = 2) +  # Smaller dots
        aes(color = interaction(Variable, Region)) +  # Different color for different variable-region combinations
        labs(title = "Time Series of Selected Variables",
             y = "Value") +
        theme_minimal() +
        theme(legend.position = "bottom") +
        scale_linetype_manual(values = c("solid", "dashed"))  # Dashed line for forecasted data
    } else if (input$visualizationType == "Scatter Plot") {
      p <- ggplot(plot_data, aes(x = Year, y = Value, color = Variable, shape = DataSource)) +
        geom_point(size = 3) +  # Dots for scatter plot
        labs(title = "Scatter Plot of Selected Variables",
             x = "Year", y = "Value") +
        theme_minimal()
    } else if (input$visualizationType == "Bar Chart") {
      p <- ggplot(plot_data, aes(x = Year, y = Value, fill = Variable)) +
        geom_bar(stat = "identity", position = "dodge") +
        labs(title = "Bar Chart of Selected Variables",
             x = "Year", y = "Value") +
        theme_minimal()
    } else if (input$visualizationType == "Heatmap") {
      p <- ggplot(plot_data, aes(x = Year, y = Variable, fill = Value)) +
        geom_tile() +
        labs(title = "Heatmap of Selected Variables",
             x = "Year", y = "Variable") +
        scale_fill_viridis_c() +
        theme_minimal()
    }
    
    ggplotly(p)
  })
  
  # Correlation plot
  output$correlationPlot <- renderPlotly({
    data <- filtered_data()
    
    if (nrow(data) == 0) {
      return(ggplotly(ggplot() + 
                      annotate("text", x = 0.5, y = 0.5, label = "Not enough data for correlation analysis") + 
                      theme_void()))
    }
    
    selected_vars <- if (input$variableType == "WASH Variables") {
      input$washVarsSelect
    } else if (input$variableType == "Disease Rates") {
      input$diseaseSelect
    } else {  # Compare Both
      c(input$washVarsSelect, input$diseaseSelect)
    }
    
    # Compute correlation between selected variables
    cors <- cor(data[selected_vars], use = "complete.obs")
    
    # Create correlation heatmap
    p <- ggplot(as.data.frame(as.table(cors)), aes(Var1, Var2, fill = Freq)) +
      geom_tile() +
      scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
      labs(title = "Correlation Between WASH Variables and Disease Rates",
           x = "Variables", y = "Variables") +
      theme_minimal()
    
    ggplotly(p)
  })
  
  # Data table output
  output$dataTable <- DT::renderDataTable({
    data <- filtered_data()
    selected_vars <- if (input$variableType == "WASH Variables") {
      input$washVarsSelect
    } else if (input$variableType == "Disease Rates") {
      input$diseaseSelect
    } else {  # Compare Both
      c(input$washVarsSelect, input$diseaseSelect)
    }
    
    DT::datatable(
      data %>% select(Region, Year, DataSource, all_of(selected_vars)),
      options = list(pageLength = 15)
    )
  })
}

# Run the application
shinyApp(ui = ui, server = server)

```
